Variable Selection in Convex Function Estimation

### 2014-04-18 tex/scam/additive_faithful.tex
lots of changes
paper update

\begin{lemma}
\label{lem:general_int_reduction}
Let $F$ be a distribution on $C=[0,1]^s$ with a positive density function $p$. Let $f: C \rightarrow \R$ be an integrable function.

Let $\ds f^*_1,...,f^*_s, \mu^* \coloneqq 
\arg\min \{ \E \big( f(X) - \sum_{k=1}^s f_k(X_k) - \mu \big)^2 \,:\, \forall k, \E f_k(X_k) = 0\}$

@@

Suppose we have the additive approximation:
\begin{equation}
f_k^*, \mu^* \coloneqq \argmin_{f_1,\ldots, f_s, \mu} \Bigl\{ 
             \E ( f(X) - \sum_{k=1}^s f_k(X_k) -\mu )^2 
         \,:\, \E f_k(X_k) = 0 \Bigr\}.
\end{equation}

@@

\begin{proof} (of Theorem~\ref{thm:convex_faithful})\\
Fix $k$. Using the result of Lemma~\ref{lem:general_int_reduction}, we need only show that for all $x_k$, $ \E[ f(X) - \sum_{k'} f_{k'}(X_{k'}) \given x_k] - \E f(X) = 0 $ implies that $f$ does not depend on coordinate $k$.\\

@@

Now we apply the first-order condition of convex functions to any two points $(x_k, \mathbf{x}_{-k})$ and $(x'_k, \mathbf{x}_{-k})$ where $x_k, x'_k \in [0,1]$:

\begin{align*}
\forall \mathbf{x}_{-k}, f(x'_k, \mathbf{x}_{-k}) & \leq f(x_k, \mathbf{x}_{-k}) 
  + f'(x_k, \mathbf{x}_{-k}) ( x'_k - x_k) \\ 
f(x'_k, \mathbf{x}_{-k}) &\leq f(x_k, \mathbf{x}_{-k}) \\
\forall \mathbf{x}_{-k}, f(x_k, \mathbf{x}_{-k}) & \leq f(x'_k, \mathbf{x}_{-k}) 
  + f'(x'_k, \mathbf{x}_{-k}) ( x_k - x'_k) \\ 
f(x_k, \mathbf{x}_{-k}) &\leq f(x'_k, \mathbf{x}_{-k})
\end{align*}

We thus have that $f(x_k, \mathbf{x}_{-k}) = f(x'_k, \mathbf{x}_{-k})$ for all $\mathbf{x}_{-k}$ and all pairs $x_k, x'_k \in (0,1)$. This proves that $f$ does not depend on $x_k$.
\end{proof}

@@

\begin{remark} 
It is difficult to prove the opposite direction of additive faithfulness, that is, if $f$ does not depend on coordinate $k$, then $f_k^*$ will be zero in the additive approximation. Consider as a conceptual example a 3D distribution over $(X_1, X_2, X_3)$; suppose $X_1, X_2$ are independent, and $f$ is only a function of $X_1, X_2$. We can then let $X_3 = f(X_1, X_2) - f^*_1(X_1) - f^*_2(X_2)$, that is, we let $X_3$ exactly capture the additive approximation error, then the best additive approximation of $f$ would have a component $f^*_3(X_3) = X_3$ even though $f$ does not depend on $X_3$. This is, of course, unlikely to occur in practice and we leave as future work analysis under reasonable assumptions that rules out such psychotic phenomenon. 
\end{remark}

>>>>

\begin{lemma}
\label{lem:general_int_reduction}
Let $F$ be a distribution on $C=[0,1]^s$ with a positive density function $p$. Let $f: C \rightarrow \R$ be an integrable function。

Let $\ds f^*_1,...,f^*_s, \mu^* \coloneqq 
\arg\min \{ \E \big( f(X) - \sum_{k=1}^s f_k(X_k) - mu \big)^2 \,:\, \forall k, \E f_k(X_k) = 0\}$

@@

If $f$ is differentiable, then $f$ depends on $k$ if $\partial_{x_k} f \neq 0$ with probability greater than 0.\\

Suppose we have the additive approximation:
\begin{equation}
\label{eqn:unconstrained_additive}
f_k^*, \mu^* \coloneqq \argmin_{f_1,\ldots, f_s, \mu} \Bigl\{ 
             \E ( f(X) - \sum_{k=1}^s f_k(X_k) -\mu )^2 
         \,:\, \E f_k(X_k) = 0 \Bigr\}.
\end{equation}

@@

\begin{proof} (of Theorem~\ref{thm:convex_faithful})\\
Fix $k$. Using the result of Lemma~\ref{lem:general_int_reduction}, we need only show that for all $x_k$, $ \E[ f(X) - \sum_{k'} f_{k'}(X_{k'}) \given x_k] - \E f(X) = 0 $ implies that $f$ does not depend on coordinate $k$, i.e., $\partial_{x_k} f(\mathbf{x}) = 0$ for all $\mathbf{x}$.\\

@@

$f$ thus does not depend on $x_k$.
% Now we apply the first-order condition of convex functions to any two points $(x_k, \mathbf{x}_{-k})$ and $(x'_k, \mathbf{x}_{-k})$ where $x_k, x'_k \in [0,1]$:

% \begin{align*}
% \forall \mathbf{x}_{-k}, f(x'_k, \mathbf{x}_{-k}) & \leq f(x_k, \mathbf{x}_{-k}) 
%   + f'(x_k, \mathbf{x}_{-k}) ( x'_k - x_k) \\ 
% f(x'_k, \mathbf{x}_{-k}) &\leq f(x_k, \mathbf{x}_{-k}) \\
% \forall \mathbf{x}_{-k}, f(x_k, \mathbf{x}_{-k}) & \leq f(x'_k, \mathbf{x}_{-k}) 
%   + f'(x'_k, \mathbf{x}_{-k}) ( x_k - x'_k) \\ 
% f(x_k, \mathbf{x}_{-k}) &\leq f(x'_k, \mathbf{x}_{-k})
% \end{align*}

% We thus have that $f(x_k, \mathbf{x}_{-k}) = f(x'_k, \mathbf{x}_{-k})$ for all $\mathbf{x}_{-k}$ and all pairs $x_k, x'_k \in (0,1)$. This proves that $f$ does not depend on $x_k$.
\end{proof}

@@

\begin{remark} 
It is difficult to prove the opposite direction of additive faithfulness, that is, if $f$ does not depend on coordinate $k$, then $f_k^*$ will be zero in the additive approximation. Consider as a conceptual example a 3D distribution over $(X_1, X_2, X_3)$; suppose $X_1, X_2$ are independent, and $f$ is only a function of $X_1, X_2$. We can then let $X_3 = f(X_1, X_2) - f^*_1(X_1) - f^*_2(X_2)$, that is, we let $X_3$ exactly capture the additive approximation error, then the best additive approximation of $f$ would have a component $f^*_3(X_3) = X_3$ even though $f$ does not depend on $X_3$. 
\end{remark}

\subsection{Convex Additive Model}

Although convex functions are additively faithful, it is difficult to estimate the optimal additive functions $f^*_k$'s (as defined in equation~\ref{eqn:unconstrained_additive}) because $f^*_k$ need not be a convex function.

Because the true regression function $f$ is convex, it is natural to consider an convex additive model:
\begin{equation}
\label{eqn:convex_additive}
\{ f^*_k \}_{k=1}^s = \arg\min \Big \{ 
    \E\big( f(X) - \sum_{k=1}^s f_k(X_k) \big)^2 \,:\, f_k \in \mathcal{C}^1, \, \E f_k(X_k) = 0 \Big \}
\end{equation}
where $\mathcal{C}^1$ is the set of univariate convex functions.

The convex functions $f^*_k$'s are not additively faithful by themselves but faithfulness can be restored by coupling the $f^*_k$'s with a set of concave functions:
\begin{equation}
\label{eqn:concave_postprocess}
g^*_k = \arg\min \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 
    \,:\, g_k \in \mh \mathcal{C}^1, \E g_k(X_k) = 0 
  \Big\}
\end{equation}

\begin{theorem}
\label{thm:acdc_faithful}
Suppose $p(\mathbf{x})$ is a positive density on $C=[0,1]^p$ and satisfies the boundary-points condition. Suppose that $f$ is convex and twice-differentiable.

Suppose that $\partial_{x_k} f$, $\partial_{x_k} p( \mathbf{x}_{-k} \given x_k )$, and $\partial_{x_k}^2 p( \mathbf{x}_{-k} \given x_k)$ are all continuous as functions on $C$.\\

Then, $f^*_k = 0$ and $g^*_k = 0$ (as defined in equation~\ref{eqn:convex_additive} and equation~\ref{eqn:concave_postprocess}) implies that $f$ does not depend on $x_k$, i.e., $\partial_{x_k} f(\mathbf{x}) = 0$ with probability 1.
\end{theorem}

Theorem~\ref{thm:acdc_faithful} suggests a two-stage screening procedure for variable selection. We first fit a sparse \emph{convex} additive model and then, for every variable marked as irrelevant in the first stage, we allow ourself to change our mind by fitting an univariate \emph{concave} function. We refer to this procedure as AC/DC (additive convex / decoupled concave) and describe it in more details in section~\ref{sec:acdc}.

\begin{proof} (of theorem~\ref{thm:acdc_faithful}) \\
Fix $k$. Let $f^*_k, g^*_k$'s be defined as equation~\ref{eqn:convex_additive} and equation~\ref{eqn:concave_postprocess}.

By definition of $f^*_k$ and $g^*_k$, we have that:
\begin{align*}
f^*_k &= \arg\min_{f_k} \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 
    \,:\, f_k \in  \mathcal{C}^1,\, \E f_k(X_k) = 0 
  \Big\} \\
g^*_k &= \arg\min_{g_k} \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 
    \,:\, g_k \in \mh \mathcal{C}^1,\, \E g_k(X_k) = 0 
  \Big\}
\end{align*}

Therefore, $f^*_k = 0$ and $f^*_k = 0$ implies that $\E[ f(X) - \sum_{k'\neq k} f^*_{k'}(X_{k'}) \given x_k] = 0$.

Now we use corollary~\ref{cor:faithfulness_extension}. We let $\phi(\mathbf{x}_{-k}) = f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'})$ and conclude that $f$ does not depend on $x_k$. 
\end{proof}

\begin{corollary} 
\label{cor:faithfulness_extension}
Suppose $p(x)$ is a positive density on $[0,1]^p$ and it satisfies the boundary-points condition.

For any function $\phi(X_{-k})$ that does not depend on $X_k$:
 \begin{align*}
f^*_k(x_k) &= \arg\min_{f_k} \mathbb{E} \Big( f(X) 
           - \phi(X_{-k}) - f_k(X_k) \Big)^2 
      = \mathbb{E}\Big[ f(X) - \phi(X_{-k}) \,|\, x_k\Big] 
\end{align*}
We have that $ f^*_k = 0 \Rightarrow \partial_{x_k} f(x) = 0$.
\end{corollary}

\begin{proof}
In the proof of theorem~\ref{thm:convex_faithful}, the only property of $r(\mathbf{x}_{-k})$ we used was the fact that $\partial_{x_k} r(\mathbf{x}_{-k}) = 0$. Therefore, the proof here is identical to that of theorem~\ref{thm:convex_faithful} except we let $\phi(\mathbf{x}_{-k})= r(\mathbf{x}_{-k}).$
\end{proof}

\begin{proposition}
\label{prop:shape_approx_univariate}
Let $C \subset \R^p$ be a compact set and let $h : C \rightarrow \R$. Let $p(x)$ be a positive density on $C$ and suppose $\E h(X) = 0$. Suppose that $\partial_{x_k} h(x)$, $\partial_{x_k} p( x \given x_k)$, and $\partial^2_{x_k} p( x \given x_k )$ are all continuous as functions on $C$. Suppose that $\partial^2_{x_k} h(x) \geq 0$.

Let 
\begin{align*}
f^*_k(x_k) &= \arg\min_{f_k} \Big\{
   \E \Big( h(X) - f_k(X_k) \Big)^2 \,:\, f_k \in \mathcal{C}^1,\, \E f_k = 0 \Big\} \\
g^*_k(x_k) &= \arg\min_{g_k} \Big \{
   \E \Big( h(X) - g_k(X_k) \Big)^2 \,:\, g_k \in \mh \mathcal{C}^1 ,\, \E g_k = 0 \Big\}
\end{align*}
be the best convex and concave univariate approximation respectively.\\

Then, $f^*_k = 0$ and $g^*_k = 0$ iff $h^*_k(x_k) \equiv \E[ h(X) \given x_k ] = 0$.
\end{proposition}

\begin{proof}

First, we will establish that $h^*_k(x_k)$ is twice differentiable and that $\partial^2_{x_k} h^*_k(x_k)$ is lower bounded.

\begin{align*}
h^*_k(x_k) &= \E[ h(X) \given x_k] \\
  &= \int_{x_{-k}} h(x) p(\mathbf{x}_{-k} \given x_k) d \mathbf{x}_{-k}\\
\partial^2_{x_k} h^*_k(x_k) 
  &= \int_{\mathbf{x}_{-k}} h''(\mathbf{x}) p(\mathbf{x}_{-k} \given x_k) +
  2 h'(x) p'(\mathbf{x}_{-k} \given x_k) + h(x) p''(\mathbf{x}_{-k} \given x_k) 
  d\mathbf{x}_{-k} 
\end{align*}

The first term $h''(x) p(\mathbf{x}_{-k} \given x_k)$ is strictly positive. By assumption, the remaining terms are continuous and hence bounded on a compact set. $\partial^2_{x_k} h^*_k(x_k)$ is therefore lower bounded. Before proceeding, we also note that because $\E h(X) = 0$, it must be that $\E h^*_k(X_k) = 0$.

Now suppose $f^*_k = 0$ and $g^*_k = 0$. Let $\sigma_k^2$ denote $\E X_k^2$. Then 
\[
\arg\min_{c \in \mathbb{R}} \mathbb{E}\Big( h(X) - c (X_k^2-\sigma_k^2) \Big)^2 = 0
\]
Since optimal $c^* = \frac{\mathbb{E} [ h(X) (X_k^2-\sigma_k^2) ]}{\mathbb{E}[ X_k^2 ]}$, 
we know $\mathbb{E} [ h(X) X_k^2 ] = 
\mathbb{E} \Big[ \mathbb{E}[ h(X) \,|\, X_k] X_k^2 \Big] = 0$.\\

Because $\partial^2_{x_k} h^*_k(x_k)$ is lower bounded, for large enough $\alpha$, $h^*_k(x_k) + \alpha (x_k^2 - \sigma_k^2)$ has a non-negative second derivative and thus is convex. Then
\[
\arg\min_{c \in \mathbb{R}} \mathbb{E}\Big( h(X) 
       - c (h^*_k(X_k) + \alpha (X_k^2 - \sigma_k^2)) \Big)^2 = 0
\]
Again, $c^* = \frac{\mathbb{E}[g(X)\big( 
            h^*_k(X_k) + \alpha (X_k^2 - \sigma_k^2) \big)]}{\mathbb{E}
        \big( h^*_k(X_k) + \alpha (X_k^2 - \sigma_k^2) \big)^2} = 0$, so
\begin{align*}
\mathbb{E}[ h(X) \big( h^*_k(X_k) + \alpha X_k^2 \big) ] &= 
\mathbb{E}[ h(X) h^*_k(X_k) ] \\
& = \mathbb{E}\Big[ \mathbb{E}[h(X) | X_k]  h^*_k(X_k) \Big] \\
& = \mathbb{E} h^*_k(X_k)^2 = 0
\end{align*}
Therefore, $h^*_k(x_k) = 0$.

\end{proof}
